---
title: "Getting started with palettome"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with palettome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

```{r setup}
library(palettome)
```

## What this package is for

When you plot clustered data — scRNA-seq clusters, cell subtypes, any
labelled embedding — the default categorical palette treats every cluster as
unrelated. But clusters usually *are* related: several T-cell subtypes belong
to one compartment, several myeloid subtypes to another. `palettome` assigns
colors that encode that structure, while keeping every individual color
overridable by hand.

The workflow is four steps:

1. `as_cluster_data()` — normalize a data frame (or a Seurat object) into a
   common structure.
2. `detect_families()` — group clusters into families, automatically or from
   a column you supply.
3. `generate_palette()` — produce the colors.
4. `cluster_colors()` / `plot_*()` / `export_*()` — use them.

Everything below runs with no UI packages installed; the interactive Shiny
app (`launch_palettome_ui()`) is entirely optional.

## Example data

The vignette uses a small synthetic embedding so it builds anywhere: three
compartments, each with a few subclusters scattered around it.

```{r example-data}
set.seed(42)

compartments <- list(
  T     = list(center = c(-5,  4), subclusters = c("T_naive", "T_mem", "T_reg", "T_cyto")),
  Mono  = list(center = c( 6,  2), subclusters = c("Mono_cl", "Mono_nc", "Mac")),
  B     = list(center = c( 0, -6), subclusters = c("B_naive", "B_mem", "Plasma"))
)

cells <- do.call(rbind, lapply(names(compartments), function(comp) {
  info <- compartments[[comp]]
  do.call(rbind, lapply(info$subclusters, function(cl) {
    offset <- stats::rnorm(2, sd = 1.6)
    n <- sample(150:400, 1)
    data.frame(
      UMAP_1  = stats::rnorm(n, info$center[1] + offset[1], 0.7),
      UMAP_2  = stats::rnorm(n, info$center[2] + offset[2], 0.7),
      cluster = cl,
      compartment = comp,
      stringsAsFactors = FALSE
    )
  }))
}))

str(cells)
```

## Step 1: build a `palettome_data` object

`as_cluster_data()` is the single entry point. Give it the coordinate columns
and the cluster column:

```{r as-cluster-data}
pdata <- as_cluster_data(
  cells,
  coord_cols  = c("UMAP_1", "UMAP_2"),
  cluster_col = "cluster"
)

pdata$coord_cols
pdata$has_family
```

`compute_centroids()` gives the per-cluster summary that the rest of the
package works from — one row per cluster rather than one per cell, which is
what keeps recoloring cheap on large datasets:

```{r centroids}
compute_centroids(pdata)
```

### The same step with a Seurat object

`as_cluster_data()` has a `Seurat` method, so you do not have to pull out
embeddings and metadata yourself. Step 1 is the only step that differs —
everything downstream sees the same `palettome_data` structure either way.

The chunk below is not run when the vignette builds (it needs a real Seurat
object), but it is the whole substitution:

```{r seurat, eval = FALSE}
# --- replaces the as_cluster_data() call above ---------------------------
# pdata <- as_cluster_data(
#   seurat_obj,
#   reduction   = "umap",            # any reduction name (Seurat::Embeddings())
#   dims        = 1:2,               # or 1:3 for a 3D embedding
#   cluster_col = "seurat_clusters", # any column of seurat_obj[[]]
#   family_col  = NULL               # optional: metadata column with an
#                                    # explicit compartment/family label
# )
#
# --- from here on, identical to the data-frame path ----------------------
# fam     <- detect_families(pdata, k = 3)
# session <- generate_palette(fam$assignment, mode = "harmonious",
#                             neighbors = fam$neighbors, seed = 1)
#
# --- and hand the colors straight back to Seurat's own plotting ----------
# Seurat::DimPlot(seurat_obj, reduction = "umap",
#                 group.by = "seurat_clusters",
#                 cols = cluster_colors(session))
```

## Step 2: detect families

With no family column present, families are found by hierarchical clustering
of the cluster centroids. `k` sets how many families to cut the tree into:

```{r detect-families}
fam <- detect_families(pdata, k = 3)
fam$method
fam$assignment
```

The dendrogram is kept on the result, so you can preview the tree and re-cut
it at a different `k` without recomputing distances:

```{r dendrogram}
plot_dendrogram(fam$dendro, k = 3)
```

```{r recut}
cut_families(fam$dendro, k = 5)
```

If you already know the compartments, pass the column instead and no
clustering happens at all — `method` becomes `"explicit"`:

```{r explicit-families}
pdata_explicit <- as_cluster_data(
  cells,
  coord_cols  = c("UMAP_1", "UMAP_2"),
  cluster_col = "cluster",
  family_col  = "compartment"
)
fam_explicit <- detect_families(pdata_explicit)
fam_explicit$method
head(fam_explicit$assignment)
```

Either way, `fam$assignment` is a plain `cluster`/`family_id` data frame you
are free to edit by hand before generating colors.

## Step 3: generate a palette

### Harmonious mode

The default. Families are laid along a single analogous hue arc with a shared
lightness ramp and muted chroma, so the whole palette reads as one coordinated
sweep rather than a bag of unrelated hues.

Passing `fam$neighbors` (the centroids, or a connectivity matrix if you
supplied one) is what makes the placement spatially aware: families adjacent
in the embedding are separated, and within a family, adjacent clusters get
far-apart shades.

```{r harmonious}
session <- generate_palette(
  fam$assignment,
  mode       = "harmonious",
  neighbors  = fam$neighbors,
  seed       = 1
)

cluster_colors(session)
```

```{r harmonious-plot}
plot_palette_static(pdata, session)
```

```{r harmonious-swatches, fig.height = 4}
plot_swatches(session)
```

### Contrast mode

Every cluster instead gets a maximally distinguishable color, chosen by
greedy max-min selection in Lab space. Family membership is still available,
carried by the family's own accent color in `session$families`:

```{r contrast}
session_contrast <- generate_palette(
  fam$assignment,
  mode      = "contrast",
  neighbors = fam$neighbors,
  seed      = 1
)

family_colors(session_contrast)
```

```{r contrast-swatches, fig.height = 4}
plot_swatches(session_contrast)
```

### Steering the look

Different seeds give genuinely different sweeps — a different hue arc
altogether, not a jitter around one look:

```{r seeds, fig.height = 4}
for (s in c(3, 11)) {
  plot_swatches(generate_palette(fam$assignment, mode = "harmonious",
                                 neighbors = fam$neighbors, seed = s))
}
```

If you want a specific gradient — brand colors, or a reference ramp — give the
anchors explicitly with `sweep_anchors` and the sweep interpolates through
them in HCL space:

```{r sweep-anchors, fig.height = 4}
session_anchored <- generate_palette(
  fam$assignment,
  mode          = "harmonious",
  sweep_anchors = c("#0B3D91", "#FF5733", "#FFD166"),
  neighbors     = fam$neighbors
)
plot_swatches(session_anchored)
```

`harmonious_style = "per_family"` switches to the older look: one hue per
family, members as graded shades of it.

```{r per-family, fig.height = 4}
plot_swatches(generate_palette(fam$assignment, mode = "harmonious",
                               harmonious_style = "per_family",
                               neighbors = fam$neighbors, seed = 1))
```

## Step 4: manual overrides

Any cluster or family color can be replaced by hand. The override is marked
with `manual_color = TRUE` and survives regeneration — a new mode, a new seed,
or an edited family assignment — until you reset it explicitly.

```{r manual}
session <- set_manual_color(session, cluster = "Plasma", color = "#2E5A87")
session$clusters[session$clusters$manual_color, c("cluster", "color", "manual_color")]
```

Regenerating now keeps that pick fixed and, because `lightness_range` and
`chroma_range` are `"auto"`, refits the remaining colors into the
lightness/chroma envelope your pick implies:

```{r regenerate}
session2 <- generate_palette(
  fam$assignment,
  session   = session,
  mode      = "contrast",
  neighbors = fam$neighbors
)
cluster_colors(session2)[["Plasma"]]
```

Families can be overridden the same way. `set_manual_color()` itself only
records the pick; regenerating then reshades that family's non-manual member
clusters around the new hue:

```{r manual-family}
session2 <- set_manual_color(session2, family = "F1", color = "#7A4E9E")
session2 <- generate_palette(fam$assignment, session = session2,
                             mode = "harmonious", neighbors = fam$neighbors)

family_colors(session2)
cluster_colors(session2, format = "data.frame")
```

`optimize_color()` snaps an arbitrary hand-picked hex to the nearest in-gamut,
aesthetically tuned lightness/chroma for its hue — useful when a raw pick is
too neon or too muddy for the surrounding palette:

```{r optimize}
optimize_color("#7B00FF", session = session2)
```

`reset_overrides()` clears them, either all at once or by name:

```{r reset}
session3 <- reset_overrides(session2, cluster = "Plasma", family = FALSE)
any(session3$clusters$manual_color)
```

## Checking the palette

`color_distance()` returns the pairwise perceptual distance matrix (CIE2000
by default) — a quick way to find the two clusters most likely to be confused:

```{r distance}
d <- color_distance(cluster_colors(session))
round(d[1:4, 1:4], 1)
```

`simulate_cvd()` maps colors through a color-vision-deficiency simulation, and
both plotting functions take a `cvd` argument so you can preview the palette
as a viewer with that deficiency would see it:

```{r cvd, fig.height = 4}
simulate_cvd(cluster_colors(session)[1:4], type = "deutan")
plot_swatches(session, cvd = "deutan")
```

## Using the colors elsewhere

`cluster_colors()` returns exactly what other plotting code wants — a named
vector keyed by cluster id:

```{r use-colors, eval = FALSE}
Seurat::DimPlot(seurat_obj, cols = cluster_colors(session))

ggplot2::ggplot(cells, ggplot2::aes(UMAP_1, UMAP_2, color = cluster)) +
  ggplot2::geom_point() +
  ggplot2::scale_color_manual(values = cluster_colors(session))
```

Or, if the copy-paste ready named vector is prefered:
```{r color-hard-code, eval=FALSE}
cluster_colors(session) |> dput()
```

The `"data.frame"` format also merges in each cluster's family id and the
family's own color, so faceting or legend code needs no extra join:

```{r colors-df}
head(cluster_colors(session, format = "data.frame"))
```

## Saving and resuming

Sessions round-trip through JSON, which is the stable contract documented in
the README — so a palette edited today can be reloaded (or read by a
non-R frontend) later:

```{r json}
path <- tempfile(fileext = ".json")
export_palette_json(session, path)

session_back <- import_palette_json(path)
identical(cluster_colors(session_back), cluster_colors(session))
```

`export_palette_csv()` writes the same tables as flat CSV when that is easier
to hand off.

```{r csv}
export_palette_csv(session, clusters_file = tempfile(fileext = ".csv"))
```

## Large datasets

Interactive editing does not need every cell drawn. `downsample_stratified()`
takes a cluster-stratified subsample — with a floor per cluster so rare
clusters stay visible — for previews; keep the full object for the final
export.

```{r downsample}
preview <- downsample_stratified(pdata, max_n = 500, min_per_cluster = 20)
nrow(preview$cells)
table(preview$cells$cluster)
```

Note that the palette itself does not change: colors come from the cluster and
family structure, not from the number of cells plotted.

## The interactive UI

Everything above is core, and depends only on `colorspace`, `farver`,
`jsonlite`, and base R. The Shiny app — drag-to-regroup families, HSL sliders,
live HCL readout, per-cluster color picking — is one optional function:

```{r ui, eval = FALSE}
launch_palettome_ui(pdata, session = session)
```

It requires `shiny`, `plotly`, `colourpicker`, and `shinyjs`, all
`Suggests`-only. If they are missing, it stops with a message naming them and
nothing else in the package is affected.
